Fast Sorting on Modern Computers

### 2005-07-27 docs/tech_report/merge.tex
reread the merge - updated results and created expected vs actual added a label to heap added a label to smallplot in report change multi mergesort to multi-mergesort

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimizations LaMarca
describes on it.

@@

Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
15\% lower instruction count than Algorithm N\footnote{This is observed on
random data, which algorithm N is not designed for. On more regular data, or
semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
is a very difficult algorithm to modify. Knuth describes it in assembly, and
making the optimizations that LaMarca describes on it is more difficult than
rewriting the algorithm more simply. The rewritten algorithm, referred to from
here on as `base mergesort', is a simple higher level version of algorithm S,
and should perform only slightly more slowly.

A disadvantage of this method is that if an array has slightly more than $2^x$
keys, an entire extra merge step would be required, which would considerably
slow the algorithm. This problem is shared by all other versions of mergesort
considered here, except algorithm N. However, since our tests use exact
powers of two as set sizes, this problem does not affect them. LaMarca,
however, used set sizes such as 4096000, which at times would suffer from this
type of problem.

@@

Overall, these three optimizations reduced the instruction count by 30\%.

\section{Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated, which can be expensive for small arrays. The address of
the auxiliary array is then offset such that the block index in the source array
is $CacheSize/2$ from the block index in the target array. The is achieved by
inverting the most significant bit of the block index in the target array.

@@

The next problem considered is that of temporal locality. Once a key is loaded
into the array, it's only used once per merge step. If the data does not fit in
half the cache, then conflict misses will occur, and no temporal reuse will
occur at all. The solution is to apply a technique used in compilers called
\n{tiling}. Using this technique, an entire cache sized portion of the array will
be fully sorted before moving on to the next set. When all the sets have been
sorted in the manner, they are all merged together as usual.

@@

To avoid this, a priority queue is used. The smallest key from each list is
added to the queue, and then put into the final target array. This reduces the
problem of the search overhead, but does not reduce the problem of
\n{thrashing}\footnote{Thrashing occurs when two arrays are mapped to the same
cache block, and the constant access of one followed by the other causes every
access to be a miss.}.  To avoid this, an entire cache line is added at a time
to the priority queue. There will still be conflict misses, but there will be no
thrashing since the keys overwritten in the cache will not be needed again.

LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
the same reasons as before. The heap is aligned, and a cache line
is put into the array each time. The smallest key is at the top of the heap,
and this is put into its final position in the array. The heap is then fixed.

@@

Algorithm N is not optimized for random data, and should not perform as well as
algorithm S. Algorithm S is expected to have a lower instruction count than base
mergesort. Neither algorithm is expected to have good cache performance,

@@

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully associative caches should not have these misses, and the results should be
correspondingly lower. Due to their design, the double mergesorts should have
better performance than their counterparts, both in level 1 misses, and overall,
though the difference is not expected to be great.

@@

As expected, algorithm N has lower instruction count than algorithm S, which
also slightly outperforms base mergesort.  The number of instructions, shown in
Figure \ref{merge instructions} is exactly as predicted, with a 50\% spike at
multi-mergesort. Level 2 misses, shown in Figure \ref{merge level 2 misses} are
also as expected, although the multi-mergesort does not show the same
improvement over tiled mergesort as in LaMarca's results.

The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
The reason for this is not clear.

As predicted, the rate of level 1 misses is greatly reduced by the double
tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
has consistently four times fewer level 1 misses than multi-mergesort, and both
double tiled sorts perform significantly better than the unoptimized sorts.

The level 2 cache results are startlingly similar to LaMarca's. The shape of the
Figure \ref{merge level 2 misses} is exactly that provided by LaMarca. The
ratio's between the each of the sorts is also the same, though the numbers are
not identical, due to different \c{int} sizes in our tests.

>>>>

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimizations LaMarca
describes on it. We attempted to contact him in order to discuss this, but
recieved no response. It is likely that the improvements were actually applied
to Knuth's algorithm S.

@@

Algorithm S is a \n{straight} merge, and works as described above.  A
disadvantage of this method is that if an array has slightly more than $2^x$
keys, an entire extra merge step would be required, which would considerably
slow the algorithm. This problem is shared by all other versions of mergesort
considered here, except algorithm N. However, since our tests use exact powers
of two as set sizes, this problem does not affect them. LaMarca, however, used
set sizes such as 4096000, which at times would suffer from this type of
problem.

Algorithm S has 15\% lower instruction count than Algorithm N\footnote{This is
observed on random data, which algorithm N is not designed for. On more regular
data, or semi-sorted data, this will probably not occur.}. Algorithm S, like
algorithm N, is a very difficult algorithm to modify. Knuth describes it in
assembly, and making the optimizations that LaMarca describes on it is more
difficult than rewriting the algorithm more simply. The rewritten algorithm,
referred to from here on as `base mergesort', is a simple higher level version
of algorithm S, and should perform only slightly more slowly.

@@

Overall, these three optimizations reduce the instruction count by 30\%.

\section{Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated. The address of
the auxiliary array is then offset such that the block index in the source array
is $CacheSize/2$ from the block index in the target array. The is achieved by
inverting the most significant bit of the block index in the target array.

@@

The next problem LaMarca considers is that of temporal locality. Once a key is
loaded into the array, it's only used once per merge step. If the data does not
fit in half the cache, then conflict misses will occur, and no temporal reuse
will occur at all. The solution is to apply a technique used in compilers called
\n{tiling}. Using this technique, an entire cache sized portion of the array
will be fully sorted before moving on to the next set. When all the sets have
been sorted in the manner, they are all merged together as usual.

@@

To avoid this, LaMarca uses a priority queue. The smallest key from each list is
added to the queue, and then put into the final target array. This reduces the
problem of the search overhead, but does not reduce the problem of
\n{thrashing}\footnote{Thrashing occurs when two arrays are mapped to the same
cache block, and the constant access of one followed by the other causes every
access to be a miss.}.  To avoid this, an entire cache line is added at a time
to the priority queue. There will still be conflict misses, but there will be no
thrashing since the keys overwritten in the cache will not be needed again.

LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
the same reasons as before (see Section \ref{cache heapsort}.). The heap is
aligned, and a cache line is put into the array each time. The smallest key is
at the top of the heap, and this is put into its final position in the array.
The heap is then fixed.

@@

Algorithm N is not optimized for random data, and should not perform as well as
algorithm S. Algorithm S is expected to have a lower instruction count than base
mergesort. Neither algorithm is expected to have good cache performance.

@@

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully associative caches should not have these misses, and the results should be
correspondingly lower. The double mergesorts should have
better performance than their counterparts in level 1 misses, though there
should be a slight increase in level 2 misses. It is expected that they should
perform slightly faster as a result.

@@

As expected, algorithm S has lower instruction count than algorithm N, and 
also slightly outperforms base mergesort.  The number of instructions, shown in
Figure \ref{merge instructions} is exactly as predicted, with a 50\% spike at
multi-mergesort.

Level 2 misses, shown in Figure \ref{merge level 2 misses} are
also as expected, although the multi-mergesort does not show the same
improvement over tiled mergesort as in LaMarca's results, as our tiled mergesort
performs better. The cache performance of algorithms N and S are poor, as
expected. They perform roughly the same as base mergesort.

It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully associative cache will perform slightly better is false. In actual fact,
the direct mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully associative cache removes the least recently used keys from the cache
to make room, while the direct mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimised
replacement policy.

The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
This shows there must be regular access patterns in the k-way merge at the end
of multi-mergesort, which are predictable using a branch history.

As predicted, the rate of level 1 misses is greatly reduced by the double
tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
has consistently four times fewer level 1 misses than multi-mergesort, and both
double tiled sorts perform significantly better than the unoptimized sorts. The
expected increase in level 2 misses did not materialise, and is in fact
negligible.

The level 2 cache results are startlingly similar to LaMarca's. The shape of the
Figure \ref{merge level 2 misses} is exactly that provided by LaMarca. The
ratio's between the each of the sorts is also the same, as each of our results
is exactly half that of LaMarca, due to different \cc{int} sizes in our tests.
